---
title: "Comparing βNTI and soil properties across the whole region"
author: "Samuel Barnett"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---


## Introduction

In this notebook, we will use the βNTI values calculated in bNTI_calculations.Rmd and see if they correlate with difference is various soil properties across all 30 soil samples. A correlation would indicate that the property is a driver of deterministic community assembly across the region.

This analysis uses the same (-2, 2) range for significance testing as in Stegen et al. 2012. This means that the following conclusison can be drawn from this data:
  βNTI > 2: Community assembly driven by variable selection
  βNTI < -2: Community assembly driven by homgenizing selection
  |βNTI| < 2: Community assembly is stochastic


### Initiate libraries
```{r, message=FALSE, warning=FALSE}
# Packages needed for analysis
library(dplyr)
library(tidyr)
library(tibble)
library(phyloseq)
library(geosphere)

# Packages needed for plotting
library(ggplot2)

```

### Import data

```{r, message=FALSE, warning=FALSE}
# Import bulk soil phyloseq data
bulk.physeq = readRDS("/home/sam/data/fullCyc2_data/bulk_soil_physeq.RDS")

## Check how many reads you have in each of the samples. This will tell you if you need to re-do anything
# Get read counts and make a new dataframe with this data
read_count = data.frame("count" = colSums(otu_table(bulk.physeq))) %>%
  rownames_to_column(var="X.Sample") %>%
  inner_join(data.frame(sample_data(bulk.physeq)), by="X.Sample") %>%
  arrange(-count) %>%
  mutate(X.Sample=factor(X.Sample, levels=X.Sample))

# Now plot read count for each sample. The horizontal line represents a 2000 read threshold
ggplot(data=read_count, aes(x=X.Sample, y=log10(count), fill=ecosystem)) +
  geom_bar(stat="identity") +
  labs(x="Sample", y="Log10(Read count)") +
  geom_hline(yintercept=log10(10000)) +
  theme(text = element_text(size=16),
        axis.text.x = element_blank())
# Everything seems to be at or above 10000 total reads

bulk.physeq
```

Now we need to rarefy the data to normalize the sequencing depth. We should also get a normalized dataset which gives relative abundance rather than readcounts.

```{r, message=FALSE, warning=FALSE}
# Rarefy to an even depth
set.seed(72)  # setting seed for reproducibility
bulk.physeq.rare = rarefy_even_depth(bulk.physeq)

# Normalize read counts (this gives relative abundance)
bulk.physeq.norm = transform_sample_counts(bulk.physeq.rare, function(x) x/sum(x))

```

Now import the βNTI data generated in bNTI_calculation.Rmd
```{r, message=FALSE, warning=FALSE}
# Import data
full.bNTI.df = read.table("/home/sam/data/fullCyc2_data/Final_data/community_assembly/full_bNTI.txt")
```

## βNTI across soil parameters.

Lets see if there is a correlation between βNTI and some of the soil parameters measured.

First we need a function for running the mantel test. This just makes things easier later with less code.

```{r, message=FALSE, warning=FALSE}

Sams.mantel.test = function(df, seed=NULL) {
  # Run mantel test to see if there is a correlation
  delta.mat = df %>%
    select(Sample_1, Sample_2, delta) %>%
    spread(Sample_2, delta)
  rownames(delta.mat) = delta.mat$Sample_1
  delta.mat$Sample_1 = NULL
  delta.mat = delta.mat[names(sort(rowSums(!is.na(delta.mat)), decreasing = F)), names(sort(colSums(!is.na(delta.mat)), decreasing = T))]
  delta.mat = as.dist(delta.mat)
  
  bNTI.mat = df %>%
    select(Sample_1, Sample_2, bNTI) %>%
    spread(Sample_2, bNTI)
  rownames(bNTI.mat) = bNTI.mat$Sample_1
  bNTI.mat$Sample_1 = NULL
  bNTI.mat = bNTI.mat[names(sort(rowSums(!is.na(bNTI.mat)), decreasing = F)), names(sort(colSums(!is.na(bNTI.mat)), decreasing = T))]
  bNTI.mat = as.dist(bNTI.mat)
  if (!(is.null(seed))){
    set.seed(seed)
  }
  mantel.res = vegan::mantel(delta.mat, bNTI.mat)
  return(mantel.res)
}
```

I also want to add the ecosystem type to the βNTI dataframe so that I can see whether we are comparing between the same or different land use strategies.

```{r, message=FALSE, warning=FALSE}
# Land use shapes
LandUse.shapes = c("agriculture" = 15, "meadow" = 16, "forest" = 17, "across" = 5)

# get habitat metadata and add it to the βNTI data
eco.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, ecosystem) %>%
  rename(Sample_1 = X.Sample, ecosystem_1 = ecosystem)
eco.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, ecosystem) %>%
  rename(Sample_2 = X.Sample, ecosystem_2 = ecosystem)

full.eco.bNTI.df = inner_join(full.bNTI.df, eco.meta1) %>%
  inner_join(eco.meta2)

```


### pH

First lets see if βNTI is correlated with the difference in pH between sites.

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=5}
# Add in pH metadata
pH.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, pH) %>%
  rename(Sample_1 = X.Sample, pH_1 = pH)
pH.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, pH) %>%
  rename(Sample_2 = X.Sample, pH_2 = pH)

full.bNTI.pH.df = inner_join(full.eco.bNTI.df, pH.meta1) %>%
  inner_join(pH.meta2) %>%
  mutate(delta = abs(pH_1-pH_2),
         crosstype = ifelse(ecosystem_1 == ecosystem_2, as.character(ecosystem_1), "across"))

# Run mantel test to see if there is a correlation
pH.mantel = Sams.mantel.test(full.bNTI.pH.df, seed=72)

# Plot
full.bNTI.pH.plot = ggplot(full.bNTI.pH.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype)) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=3.25, y=12.5, label=paste("r= ", round(pH.mantel$statistic, 3), "\n", "p= ", round(pH.mantel$signif, 3), sep="")) +
  labs(x="∆ pH", y="βNTI") +
  theme(legend.position = "none")

full.bNTI.pH.plot

```

There is a significant positive relationship between βNTI and difference in pH between sites. For the most part, sites with very similar pHs have a βNTI < -2 suggesting homogeneous selection pressure, while sites with very dissimilar pHs have a βNTI > 2 suggesting varaible selection.

### Soil organic matter

Now lets see if βNTI is correlated with the difference in soil organic matter (SOM) between sites.

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=5}
# Add in SOM metadata
SOC.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, organic_content_perc) %>%
  rename(Sample_1 = X.Sample, SOC_1 = organic_content_perc)
SOC.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, organic_content_perc) %>%
  rename(Sample_2 = X.Sample, SOC_2 = organic_content_perc)

full.bNTI.SOC.df = inner_join(full.eco.bNTI.df, SOC.meta1) %>%
  inner_join(SOC.meta2) %>%
  mutate(delta = abs(SOC_1-SOC_2),
         crosstype = ifelse(ecosystem_1 == ecosystem_2, as.character(ecosystem_1), "across"))

# Run mantel test to see if there is a correlation
SOC.mantel = Sams.mantel.test(full.bNTI.SOC.df, seed=72)

# Plot
full.bNTI.SOC.plot = ggplot(full.bNTI.SOC.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype)) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=0.1375, y=12.5, label=paste("r= ", round(SOC.mantel$statistic, 3), "\n", "p= ", round(SOC.mantel$signif, 3), sep="")) +
  labs(x="∆ %SOC", y="βNTI") +
  theme(legend.position = "none")
full.bNTI.SOC.plot

```


### C:N ratio

Now lets see if βNTI is correlated with the difference in C:N ratio between sites.

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=5}
# Add in C:N metadata
CN.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, percent_N, percent_C) %>%
  rename(Sample_1 = X.Sample) %>%
  mutate(CN_1 = percent_C/percent_N) %>%
  select(-percent_N, -percent_C)
CN.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, percent_N, percent_C) %>%
  rename(Sample_2 = X.Sample) %>%
  mutate(CN_2 = percent_C/percent_N) %>%
  select(-percent_N, -percent_C)

full.bNTI.CN.df = inner_join(full.eco.bNTI.df, CN.meta1) %>%
  inner_join(CN.meta2) %>%
  mutate(delta = abs(CN_1-CN_2),
         crosstype = ifelse(ecosystem_1 == ecosystem_2, as.character(ecosystem_1), "across"))

# Run mantel test to see if there is a correlation
CN.mantel = Sams.mantel.test(full.bNTI.CN.df, seed=72)


# Plot
full.bNTI.CN.plot = ggplot(full.bNTI.CN.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype)) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=9.3, y=12.5, label=paste("r= ", round(CN.mantel$statistic, 3), "\n", "p= ", round(CN.mantel$signif, 3), sep="")) +
  labs(x="∆ C:N", y="βNTI") +
  theme(legend.position = "none")
full.bNTI.CN.plot

```

There is a significant relationship between βNTI and delta C:N ratio.


### Percent sand

Now lets see if βNTI is correlated with the difference in percent sand between sites.

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=5}
# Add in % sand metadata
sand.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, sand__perc) %>%
  rename(Sample_1 = X.Sample, Sand1 = sand__perc)
sand.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, sand__perc) %>%
  rename(Sample_2 = X.Sample, Sand2 = sand__perc)

full.bNTI.sand.df = inner_join(full.eco.bNTI.df, sand.meta1) %>%
  inner_join(sand.meta2) %>%
  mutate(delta = abs(Sand1-Sand2),
         crosstype = ifelse(ecosystem_1 == ecosystem_2, as.character(ecosystem_1), "across"))

# Run mantel test to see if there is a correlation
sand.mantel = Sams.mantel.test(full.bNTI.sand.df, seed=72)


# Plot
full.bNTI.sand.plot = ggplot(full.bNTI.sand.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype)) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=65, y=12.5, label=paste("r= ", round(sand.mantel$statistic, 3), "\n", "p= ", round(sand.mantel$signif, 3), sep="")) +
  labs(x="∆ sand", y="βNTI") +
  theme(legend.position = "none")

full.bNTI.sand.plot
```

There is no signficiant correlaton between βNTI and delta percent sand

### Geographic distance

Now lets see if βNTI is correlated with the geographic distance between sites.

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=5}
# Get geographic distances between sites
geodist.meta1 = data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, longitude, latitude) %>%
  rename(Sample_1 = X.Sample, longitude1 = longitude, latitude1 = latitude)
geodist.meta2 = data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, longitude, latitude) %>%
  rename(Sample_2 = X.Sample, longitude2 = longitude, latitude2 = latitude)

full.bNTI.geodist.df = inner_join(full.eco.bNTI.df, geodist.meta1) %>%
  inner_join(geodist.meta2) %>%
  mutate(crosstype = ifelse(ecosystem_1 == ecosystem_2, as.character(ecosystem_1), "across"))

Coord1.mat = as.matrix(full.bNTI.geodist.df %>% select(longitude1, latitude1))
Coord2.mat = as.matrix(full.bNTI.geodist.df %>% select(longitude2, latitude2))

full.bNTI.geodist.df$delta = distHaversine(Coord1.mat, Coord2.mat)/1000

# Run mantel test to see if there is a correlation
geodist.mantel = Sams.mantel.test(full.bNTI.geodist.df, seed=72)


# Plot
full.bNTI.geodist.plot = ggplot(full.bNTI.geodist.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype)) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=42.5, y=12.5, label=paste("r= ", round(geodist.mantel$statistic, 3), "\n", "p= ", round(geodist.mantel$signif, 3), sep="")) +
  labs(x="Geographic distance (km)", y="βNTI") +
  theme(legend.position = "none")

full.bNTI.geodist.plot
```

There is a slight but significant correlation between βNTI and geographic distance when including all pairwise site combinations.

### Plot all together for publication

```{r, message=FALSE, warning=FALSE, fig.height=6.61417, fig.width=6.61417}
# pH
full.bNTI.pH.plot = ggplot(full.bNTI.pH.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype), size=3) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=3.5*.9, y=12.5, size=4,
           label=paste("r = ", round(pH.mantel$statistic, 3), "\n", "p = ", round(pH.mantel$signif, 3), sep="")) +
  labs(x="∆ pH", y="βNTI") +
  lims(x=c(0, 3.5), y=c(-10, 14)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size=12),
        axis.title= element_text(size=14),
        plot.margin = unit(c(7,1,1,5), "mm"))

# SOM
full.bNTI.SOC.plot = ggplot(full.bNTI.SOC.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype), size=3) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=0.165*.9, y=12.5, size=4, 
           label=paste("r = ", round(SOC.mantel$statistic, 3), "\n", "p = ", round(SOC.mantel$signif, 3), sep="")) +
  labs(x="∆ SOM", y="βNTI") +
  lims(x=c(0, 0.165), y=c(-10, 14)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size=12),
        axis.title= element_text(size=14),
        plot.margin = unit(c(7,1,1,5), "mm"))

# C:N ratio
full.bNTI.CN.plot = ggplot(full.bNTI.CN.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype), size=3) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=10.5*.9, y=12.5, size=4,
           label=paste("r = ", round(CN.mantel$statistic, 3), "\n", "p = ", round(CN.mantel$signif, 3), sep="")) +
  labs(x="∆ C:N", y="βNTI") +
  lims(x=c(0, 10.5), y=c(-10, 14)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size=12),
        axis.title= element_text(size=14),
        plot.margin = unit(c(7,1,1,5), "mm"))

# Geographic distance
full.bNTI.geodist.plot = ggplot(full.bNTI.geodist.df, aes(x=delta, y=bNTI)) +
  geom_point(aes(shape=crosstype), size=3) +
  scale_shape_manual(values=LandUse.shapes) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  annotate("text", x=49*.9, y=12.5, size=4,
           label=paste("r = ", round(geodist.mantel$statistic, 3), "\n", "p = ", round(geodist.mantel$signif, 3), sep="")) +
  labs(x="Geographic distance (km)", y="βNTI") +
  lims(x=c(0, 49), y=c(-10, 14)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size=12),
        axis.title= element_text(size=14),
        plot.margin = unit(c(7,1,1,5), "mm"))



bNTI.chem.plot = cowplot::plot_grid(full.bNTI.pH.plot, full.bNTI.SOC.plot, full.bNTI.CN.plot, full.bNTI.geodist.plot, labels=c("A", "B", "C", "D"))

#ggsave("bNTI_chem.tiff", plot=bNTI.chem.plot, device="tiff", 
#       path="/home/sam/notebooks/fullCyc2/figures/community_assembly_MS/",
#       width=168, height=168, units="mm", dpi=600)

bNTI.chem.plot

```


```{r}
sessionInfo()
```




