---
title: "Relative influence of sochastic and deterministic processes on community turnover"
author: "Samuel Barnett"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    toc: true
    toc_depth: 2
    html_preview: false
---


## Introduction

In the following analysis we will be using both βNTI and RCbray to examine the processes driving community assembly across land use regime and pH groups. βNTI values were calculated in bNTI_calculations.Rmd and RCbray values are calculated here

This analysis values for significance testing as in Stegen et al. 2012 and Stegen et al. 2013. This means that the following conclusison can be drawn from this data:

    βNTI > 2: Community assembly driven by variable selection
    βNTI < -2: Community assembly driven by homgenizing selection
    |βNTI| < 2 & RCbray > 0.95: Community assembly driven by dispersal limitation
    |βNTI| < 2 & RCbray < -0.95: Community assembly driven by homogenizing dispersal
    |βNTI| < 2 & |RCbray| < 0.95: Community assembly is due to drift alone (or weak selection)
  
Please note that some of these cell have eval=FALSE in the rmarkdown version so that they wont run when knitted. This was done because they take a really long time and are better run in terminal.

### Initiate libraries
```{r, message=FALSE, warning=FALSE}
# Packages needed for analysis
library(dplyr)
library(tibble)
library(phyloseq)
library(ape)
library(vegan)
library(parallel)

# Packages needed for plotting
library(ggplot2)

```

### Import data

```{r, message=FALSE, warning=FALSE}
# Import bulk soil phyloseq data
bulk.physeq = readRDS("/home/sam/data/fullCyc2_data/bulk_soil_physeq.RDS")

## Check how many reads you have in each of the samples. This will tell you if you need to re-do anything
# Get read counts and make a new dataframe with this data
read_count = data.frame("count" = colSums(otu_table(bulk.physeq))) %>%
  rownames_to_column(var="X.Sample") %>%
  inner_join(data.frame(sample_data(bulk.physeq)), by="X.Sample") %>%
  arrange(-count) %>%
  mutate(X.Sample=factor(X.Sample, levels=X.Sample))

# Now plot read count for each sample. The horizontal line represents a 2000 read threshold
ggplot(data=read_count, aes(x=X.Sample, y=log10(count), fill=ecosystem)) +
  geom_bar(stat="identity") +
  labs(x="Sample", y="Log10(Read count)") +
  geom_hline(yintercept=log10(10000)) +
  theme(text = element_text(size=16),
        axis.text.x = element_blank())
# Everything seems to be at or above 10000 total reads

bulk.physeq
```

Now we need to rarefy the data to normalize the sequencing depth. We should also get a normalized dataset which gives relative abundance rather than readcounts.

```{r, message=FALSE, warning=FALSE}
# Rarefy to an even depth
set.seed(72)  # setting seed for reproducibility
bulk.physeq.rare = rarefy_even_depth(bulk.physeq)

# Normalize read counts (this gives relative abundance)
bulk.physeq.norm = transform_sample_counts(bulk.physeq.rare, function(x) x/sum(x))

```

Now import the βNTI data generated in bNTI_calculation.Rmd

```{r, message=FALSE, warning=FALSE}
# Import data
full.bNTI.df = read.table("/home/sam/data/fullCyc2_data/Final_data/community_assembly/full_bNTI.txt")
```

## Calculating RCbray

First we need a function to calculate RCbray.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Function for calculating the distances in the null communities
RCbray_null_func <- function(i, freq.abd.df, alpha1, alpha2, N){
  # Get simulated communities and distance
  ## initally select OTUs weighted by their frequency. The number of OTUs selected should equal the richness of the samples.
  simcom1 = data.frame(table(sample(freq.abd.df$OTU, size=alpha1, replace=F, prob=freq.abd.df$freq)), stringsAsFactors = F)
  colnames(simcom1) = c("OTU","simcom1")
  simcom1$OTU = as.character(simcom1$OTU)
  simcom1 = inner_join(simcom1, freq.abd.df, by="OTU")
  simcom2 = data.frame(table(sample(freq.abd.df$OTU, size=alpha2, replace=F, prob=freq.abd.df$freq)), stringsAsFactors = F)
  colnames(simcom2) = c("OTU","simcom2")
  simcom2$OTU = as.character(simcom2$OTU)
  simcom2 = inner_join(simcom2, freq.abd.df, by="OTU")

  ## Now recruit OTUs based on their abundance in the metacommunity
  simcom1.abd = data.frame(table(sample(simcom1$OTU, size=N-alpha1, replace=T, prob=simcom1$p)), stringsAsFactors = F)
  colnames(simcom1.abd) = c("OTU","simcom1.abd")
  simcom1.abd$OTU = as.character(simcom1.abd$OTU)
  simcom1 = full_join(simcom1, simcom1.abd, by="OTU") %>%
    mutate(simcom1.abd = ifelse(is.na(simcom1.abd), 1, simcom1.abd)) %>%
    select(OTU, simcom1.abd)
  
  simcom2.abd = data.frame(table(sample(simcom2$OTU, size=N-alpha2, replace=T, prob=simcom2$p)), stringsAsFactors = F)
  colnames(simcom2.abd) = c("OTU","simcom2.abd")
  simcom2.abd$OTU = as.character(simcom2.abd$OTU)
  simcom2 = full_join(simcom2, simcom2.abd, by="OTU") %>%
    mutate(simcom2.abd = ifelse(is.na(simcom2.abd), 1, simcom2.abd)) %>%
    select(OTU, simcom2.abd)

  
  simcom = full_join(simcom1, simcom2, by="OTU")
  simcom[is.na(simcom)] = 0
  rownames(simcom) = simcom$OTU
  simcom$OTU = NULL
  
  null.dist = vegdist(t(simcom), method="bray")[1]
  return(null.dist)
}

# Main function for calculating RCbray
Calc_RCbray <- function(physeq, reps, nproc){
  # Get OTU table from phyloseq object
  otu.table = otu_table(physeq)
  
  # Get alpha diversity for each sample
  otu.PA.table = otu.table
  otu.PA.table[otu.PA.table > 0] = 1
  alpha.df = data.frame(Sample_ID = colnames(otu.PA.table), OTU.n = colSums(otu.PA.table), stringsAsFactors = F)
  
  # Get beta diversity matrix
  beta.table = as.matrix(vegdist(t(otu.PA.table), method="bray", diag=TRUE, upper=TRUE))
  
  ## Get metacommunity
  # Calculate the number of individuals in the meta community (Average read depth)
  N <- mean(apply(t(otu.table), 1, sum))
  
  # Calculate the average relative abundance of each taxa across communities
  p.m <- apply(t(otu.table), 2, mean)
  p.m <- p.m[p.m != 0]
  p <- p.m/N
  
  # Calculate the occurrence frequency of each taxa across communities
  otu.table.bi <- 1*(t(otu.table)>0)
  freq <- apply(otu.table.bi, 2, mean)
  freq <- freq[freq != 0]
  
  # Combine
  freq.abd.df = data.frame(p=p, freq=freq) %>%
    tibble::rownames_to_column(var="OTU") %>%
    filter(p != 0, freq != 0) %>%
    arrange(p)
  
  # For each pair of samples run the RCbray analysis
  comps = combn(alpha.df$Sample_ID, m=2, simplify = F)
  RCb.df = data.frame(Site1 = character(), Site2 = character(), RCb = numeric(), stringsAsFactors = F)
  for (j in seq(1, length(comps))){
    sam = comps[[j]]
    alpha1 = alpha.df[alpha.df$Sample_ID == sam[1],]$OTU.n
    alpha2 = alpha.df[alpha.df$Sample_ID == sam[2],]$OTU.n
    # Permute "reps" many times
    rep.list = seq(1, reps)
    null.list = mclapply(rep.list, RCbray_null_func, freq.abd.df=freq.abd.df, alpha1=alpha1, alpha2=alpha2, N=N, mc.cores=nproc)
    
    RCb = (length(null.list[null.list > beta.table[sam[1], sam[2]]]) + (0.5*length(null.list[null.list == beta.table[sam[1], sam[2]]])))/reps
    RCb = (RCb - 0.5)*2
    
    RCb.df = rbind(RCb.df, data.frame(Site1=sam[1], Site2=sam[2], RCb=RCb, stringsAsFactors = F))
  }
  
  RCb.df
  return(RCb.df)
}

```

Now calculate the RCbray for 999 bootstrap replicates for each site pair. This should only be run once (eval=FALSE).

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# Run the RCbray function
RCb.df = Calc_RCbray(bulk.physeq.rare, 999, 20)
write.table(RCb.df, file="/home/sam/fullCyc2_data/Final_data/community_assembly/full.RCb.df.txt", quote = F, sep = "\t")

writeLines("\n\n\n\n\nDone!!!\n\n\n\n\n")

```

## Merging βNTI and RCbray data

Now I'll import both the βNTI data and the newly generated RCbray data. Then add in land use categories.

```{r, message=FALSE, warning=FALSE}

# bNTI data
# Import bNTI data
bNTI.df = read.table("/home/sam/data/fullCyc2_data/Final_data/community_assembly/full_bNTI.txt")

# RCbray data
RCb.df = read.table(file="/home/sam/data/fullCyc2_data/Final_data/community_assembly/full.RCb.df.txt") %>%
  mutate(Sample_1 = Site2, Sample_2 = Site1)

# Get habitat metadata and add it to the βNTI then merge with the RCbray dataset
eco.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, ecosystem) %>%
  rename(Sample_1 = X.Sample, ecosystem_1 = ecosystem)
eco.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, ecosystem) %>%
  rename(Sample_2 = X.Sample, ecosystem_2 = ecosystem)

bNTI.df = inner_join(bNTI.df, eco.meta1) %>%
  inner_join(eco.meta2)
  

# Merge both datasets
turnover.df = inner_join(bNTI.df, RCb.df)
head(turnover.df)

```

## Community assembly processes across land use regime

Now lets combine the RCbray results with the results from the bNTI analysis. This will tell us the relative influcence of both deterministic (bNTI) and stochastic (RCbray) processes on community turnover in each land use regime.

```{r, fig.height=3.14961, fig.width=6.61417, message=FALSE, warning=FALSE, echo=FALSE}
# Make bNTI figure
within.bNTI.df = bNTI.df %>%
  filter(ecosystem_1 == ecosystem_2) %>%
  mutate(ecosystem = ecosystem_1)

ecosystem.conv = data.frame(ecosystem = c("agriculture", "meadow", "forest"), ecosystem2 = c("Cropland", "Old-field", "Forest"))
within.bNTI.df = left_join(within.bNTI.df, ecosystem.conv)
within.bNTI.df$ecosystem2 = factor(within.bNTI.df$ecosystem2, levels = c("Cropland", "Old-field", "Forest"))

within.bNTI.df$ecosystem = factor(within.bNTI.df$ecosystem, levels=c("agriculture", "meadow", "forest"))


eco.bNTI.plot = ggplot(within.bNTI.df, aes(x=ecosystem2, y=bNTI)) +
  geom_boxplot(outlier.shape=1) +
  geom_hline(yintercept = 2, linetype=2, size=0.5) +
  geom_hline(yintercept = -2, linetype=2, size=0.5) +
  labs(x="Land use", y="βNTI") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size=12),
        axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_text(size=14))

# Make the community analysis plot
eco.turnover.df = turnover.df %>%
  filter(ecosystem_1 == ecosystem_2) %>%
  mutate(ecosystem = ecosystem_1)

eco.turnover.df = left_join(eco.turnover.df, ecosystem.conv)
eco.turnover.df$ecosystem2 = factor(eco.turnover.df$ecosystem2, levels = c("Cropland", "Old-field", "Forest"))

eco.turnover.df$ecosystem = factor(eco.turnover.df$ecosystem, levels=c("agriculture", "meadow", "forest"))

## Calculate the relative influence of each process
eco.turnover.df = eco.turnover.df %>%
  mutate(process = ifelse(abs(bNTI) < 2,
                          ifelse(abs(RCb) < 0.95, "Drift",
                                 ifelse(RCb >= 0.95, "Dispersal Limited",
                                        ifelse(RCb <= -0.95, "Homogenizing Dispersal", "ERROR"))),
                          ifelse(bNTI >= 2, "Variable Selection",
                                 ifelse(bNTI <= -2, "Homogeneous Selection", "ERROR"))))
eco.turnover.df$process = factor(eco.turnover.df$process, levels = c("Drift", 
                                                             "Dispersal Limited", "Homogenizing Dispersal", 
                                                             "Variable Selection", "Homogeneous Selection"))

sum.eco.turnover.df = eco.turnover.df %>%
  group_by(ecosystem2, process) %>%
  dplyr::summarize(n_sites = n(),
                   perc=(n()/45)*100) %>%
  as.data.frame

sum.eco.turnover.df

## Fix the look of some of the variable names for the figure. This is purely asthetic
sum.eco.turnover.mod.df = sum.eco.turnover.df %>%
  mutate(process = gsub(" S", "\ns", process)) %>%
  mutate(process = gsub(" D", "\nd", process)) %>%
  mutate(process = gsub("Drift", "Drift alone\n ", process))
sum.eco.turnover.mod.df$process = factor(sum.eco.turnover.mod.df$process, 
                                        levels = c("Drift alone\n ",
                                                   "Homogenizing\ndispersal", 
                                                   "Variable\nselection", 
                                                   "Homogeneous\nselection"))

eco.turnover.plot = ggplot(sum.eco.turnover.mod.df, aes(x=ecosystem2, y=perc, fill=process)) +
  geom_bar(stat="identity", color="black") +
  scale_fill_manual(values = c("white", "grey75", "grey50", "black")) + 
  labs(x="Land use", y="Percent of site pairs", fill="Process") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=12),
        axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_text(size=14),
        legend.key.size = unit(10, "mm"),
        legend.text = element_text(size=12),
        legend.title = element_text(size=14))

# Merge the plots
eco.plot = cowplot::plot_grid(eco.bNTI.plot, eco.turnover.plot, 
                   rel_widths=c(0.6, 1), labels=c("A", "B"))
eco.plot

#ggsave("alt_land_use_assembly_processes.tiff", plot=eco.plot, device="tiff", 
#       path="/home/sam/notebooks/fullCyc2/figures/community_assembly_MS/",
#       width=168, height=80, units="mm", dpi=600)

```

## Community assembly processes across pH group

Now lets look at the relative influcence of both deterministic (bNTI) and stochastic (RCbray) processes on community turnover in each pH group. Again acidic soils have a pH ≤ 5.5 and pH neutral soils have 5.5 < pH < 8.5. As we found in bNTI_across_pH_groups.Rmd there is an interaction between pH group and land use regime so we want to add that to the figure too.

```{r, fig.height=3.14961, fig.width=6.61417, message=FALSE, warning=FALSE, echo=FALSE}
# Get pH metadata for each site
pH.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, pH) %>%
  mutate(pH_group_1 = ifelse(pH <= 5.5, "Acidic", "Neutral")) %>%
  rename(Sample_1 = X.Sample, pH_1 = pH)
pH.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, pH) %>%
  mutate(pH_group_2 = ifelse(pH <= 5.5, "Acidic", "Neutral")) %>%
  rename(Sample_2 = X.Sample, pH_2 = pH)

# Make bNTI figure
pH.bNTI.df = inner_join(bNTI.df, pH.meta1) %>%
  inner_join(pH.meta2) %>%
  filter(pH_group_1 == pH_group_2) %>%
  mutate(pH_group = pH_group_1)

pH.bNTI.df$pH_group = factor(pH.bNTI.df$pH_group, levels=c("Acidic", "Neutral"))

pH.bNTI.m.df = pH.bNTI.df %>%
  mutate(ecosystem = ifelse(ecosystem_1 == ecosystem_2, as.character(ecosystem_1), "across"))

pH.bNTI.plot = ggplot(pH.bNTI.m.df, aes(x=pH_group, y=bNTI)) +
  geom_boxplot(outlier.shape=1) +
  geom_hline(yintercept = 2, linetype=2, size=0.5) +
  geom_hline(yintercept = -2, linetype=2, size=0.5) +
  labs(x="pH group", y="βNTI") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size=12),
        axis.title = element_text(size=14))

# Make the community analysis plot
pH.turnover.df = inner_join(turnover.df, pH.meta1) %>%
  inner_join(pH.meta2) %>%
  filter(pH_group_1 == pH_group_2) %>%
  mutate(pH_group = pH_group_1)

## Calculate relative influence of each process
pH.turnover.df = pH.turnover.df %>%
  mutate(process = ifelse(abs(bNTI) < 2,
                          ifelse(abs(RCb) < 0.95, "Drift",
                                 ifelse(RCb >= 0.95, "Dispersal Limited",
                                        ifelse(RCb <= -0.95, "Homogenizing Dispersal", "ERROR"))),
                          ifelse(bNTI >= 2, "Variable Selection",
                                 ifelse(bNTI <= -2, "Homogeneous Selection", "ERROR"))))
pH.turnover.df$process = factor(pH.turnover.df$process, levels = c("Drift", 
                                                                   "Dispersal Limited", "Homogenizing Dispersal", 
                                                                   "Variable Selection", "Homogeneous Selection"))
sum.pH.turnover.df = pH.turnover.df %>%
  group_by(pH_group, process) %>%
  dplyr::summarize(n_sites = n(),
                   perc=(n()/105)*100) %>%
  as.data.frame

sum.pH.turnover.df

## Fix the look of some of the variable names for the figure. This is purely asthetic
sum.pH.turnover.mod.df = sum.pH.turnover.df %>%
  mutate(process = gsub(" S", "\ns", process)) %>%
  mutate(process = gsub(" D", "\nd", process)) %>%
  mutate(process = gsub("Drift", "Drift alone\n ", process))
sum.pH.turnover.mod.df$process = factor(sum.pH.turnover.mod.df$process, 
                                        levels = c("Drift alone\n ",
                                                   "Homogenizing\ndispersal", 
                                                   "Variable\nselection", 
                                                   "Homogeneous\nselection"))

pH.turnover.plot = ggplot(sum.pH.turnover.mod.df, aes(x=pH_group, y=perc, fill=process)) +
  geom_bar(stat="identity", color="black") +
  scale_fill_manual(values = c("white", "grey75", "grey50", "black")) + 
  labs(x="pH group", y="Percent of site pairs", fill="Process") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=12),
        axis.title = element_text(size=14),
        legend.text = element_text(size=12),
        legend.title = element_text(size=14),
        legend.key.size = unit(10, "mm"))


pH.plot = cowplot::plot_grid(pH.bNTI.plot, pH.turnover.plot, 
                   rel_widths=c(0.6, 1), labels=c("A", "B"))
pH.plot

```

Now add in the plot also generated in bNTI_across_pH_groups.Rmd showing the interaction between pH group and land use on bNTI.

```{r, fig.height=6.61417, fig.width=6.61417, message=FALSE, warning=FALSE, echo=FALSE}

# get habitat metadata and add it to the βNTI data
eco.meta1=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, ecosystem, location) %>%
  rename(Sample_1 = X.Sample, ecosystem_1 = ecosystem, location_1 = location)
eco.meta2=data.frame(sample_data(bulk.physeq.rare)) %>%
  select(X.Sample, ecosystem, location) %>%
  rename(Sample_2 = X.Sample, ecosystem_2 = ecosystem, location_2 = location)

eco.pH.bNTI.df = inner_join(pH.bNTI.df, eco.meta1) %>%
  inner_join(eco.meta2) %>%
  mutate(LandUse_1 = ifelse(ecosystem_1 == "agriculture" | ecosystem_2 == "agriculture", "Cropland",
                            ifelse(ecosystem_1 == "meadow" | ecosystem_2 == "meadow", "Old-field", "Forest"))) %>%
  mutate(LandUse_2 = ifelse(ecosystem_1 == ecosystem_2, LandUse_1,
                            ifelse(LandUse_1 == "Cropland",
                                   ifelse(ecosystem_1 == "meadow" | ecosystem_2 == "meadow", "Old-field", "Forest"),
                                   "Forest"))) %>%
  mutate(LandUse_cross = paste(LandUse_1, LandUse_2, sep=" vs. "),
         Location_diff = ifelse(location_1 == location_2, "Same location", "Different location"),
         LandUse_diff = ifelse(LandUse_1 == LandUse_2, "Same land use", "Different land use")) %>%
  arrange(ecosystem_1, ecosystem_2)
eco.pH.bNTI.df$LandUse_cross = factor(eco.pH.bNTI.df$LandUse_cross, levels = c("Cropland vs. Cropland", "Old-field vs. Old-field", "Forest vs. Forest",
                                                                               "Cropland vs. Old-field", "Cropland vs. Forest", "Old-field vs. Forest"))

# Plot the βNTI data for each pH_group
eco.pH.bNTI.plot = ggplot(eco.pH.bNTI.df, aes(x=LandUse_cross, y=bNTI)) +
  geom_boxplot(aes(fill=LandUse_diff), outlier.shape=1) +
  #geom_jitter(aes(shape=Location_diff)) +
  scale_fill_manual(values=c("lightgrey", "white")) +
  scale_shape_manual(values=c(1, 16)) +
  geom_hline(yintercept = 2, linetype=2) +
  geom_hline(yintercept = -2, linetype=2) +
  labs(x="Land use of site pairs", y="βNTI", fill="Site pair category", shape="Site locations") +
  facet_grid(~pH_group) +
  theme_bw() +
  theme(axis.text = element_text(size=12),
        axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_text(size=14),
        legend.text = element_text(size=12),
        legend.title = element_text(size=14),
        strip.text.x = element_text(size=14),
        legend.position="none",
        plot.margin = unit(c(5,1,1,10), "mm"))

inter.pH.plot = cowplot::plot_grid(pH.plot, eco.pH.bNTI.plot, labels=c("", "C"), nrow=2)
inter.pH.plot

#ggsave("pH_assembly_processes_wInter.tiff", plot=alt.inter.pH.plot, device="tiff", 
#       path="/home/sam/notebooks/fullCyc2/figures/community_assembly_MS/",
#       width=168, height=168, units="mm", dpi=600)
```

## Session Info

```{r}
sessionInfo()
```
